Pushdown Control-Flow Analysis via Refunctionalization (Functional Pearl)

### 2018-03-11 paper.tex
update

  Remarkably, refunctionalization sequentializes the abstract evaluation.
  Following the idea that evaluation contexts are defunctionalized 
  continuations, we reveal how precise call/return match in control-flow 
  analysis can be obtained by properly refunctionalizing a small-step abstract interpreter.
  As well as we explain how it been lost by defunctionalizing the higher-order
  continuations into an first-order data type.

@@

\subsection{Pushdown Machine}

AAM with unbounded stack

@@

In last section, we showed that by keeping an unabstracted stack in the state 
space we can recover the precise call/return match. In this section, we start
our first transformation: \textit{fusing} an AAM with unbounded stack.

In fact, later we will eliminate the stack and transform it to continuations
represented by higher-order functions.
Because of transforming to higher-order functions forces us to sequentialize the execution order,
it does not matter what kind of AAM we start from. 
We choose to start from AAM with unbounded stack just because TODO.

\begin{lstlisting}
def drive_step(todo: List[State], seen: Set[State]): Set[State] = todo match {
  case Nil => seen
  case hd::tl if seen.contains(hd) => drive_step(tl, seen)
  case hd::tl =>
    val new_time = tick(hd)
    val new_states = hd match {
      case State(Let(x, App(f, ae), e), env, bstore, konts, time) 
        if isAtomic(ae) =>
        for (Clos(Lam(v, body), env_c) <- atomicEval(f, env, bstore).toList) yield {
          val frame = Frame(x, e, env)
          val baddr = allocBind(v, new_time)
          val new_env = env_c + (v -> baddr)
          val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
          State(body, new_env, new_store, frame::konts, new_time)
        }
      case State(ae, env, bstore, konts, time) if isAtomic(ae) =>
        konts match {
          case Nil => List()
          case Frame(x, e, f_env)::konts =>
            val baddr = allocBind(x, new_time)
            val new_env = f_env + (x -> baddr)
            val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
            List(State(e, new_env, new_store, konts, new_time))
        }
    }
    drive_step(new_states ++ tl, seen + hd)
}

@@

\begin{lstlisting}
def aval(state: State, todo: List[State], seen: Set[State]): Set[State] = { 
  val newTime = tick(state)
  state match {
    case State(Let(x, App(f, ae), e), env, bstore, konts, time) if isAtomic(ae) =>
      val newStates = for (Clos(Lam(v, body), c_env) <- aeval(f, env, bstore).toList) yield {
        val frame = Frame(x, e, env)
        val baddr = allocBind(v, newTime)
        val newEnv = c_env + (v -> baddr)
        val newStore = bstore.update(baddr, aeval(ae, env, bstore))
        State(body, newEnv, newStore, frame::konts, newTime)
      }   
      collect(newStates ++ todo, seen + state)
    case State(ae, env, bstore, konts, time) if isAtomic(ae) =>
      continue(konts, aeval(ae, env, bstore), bstore, newTime, todo, seen + state)
  }   
}

def collect(todo: List[State], seen: Set[State]): Set[State] = { 
  todo match {
    case Nil => seen
    case hd::tl if seen.contains(hd) => collect(tl, seen)
    case hd::tl => aval(hd, tl, seen)
  }   
}

@@

@@

\subsubsection{First Try}

@@

def aval(e: Expr, env: Env, store: BStore, time: Time, cont: Cont, rcont: MCont): Set[VS] = {
  val new_time = (e::time).take(k)
  e match {
    case Let(x, App(f, ae), e) =>
      val closures = aeval(f, env, store).toList.asInstanceOf[List[Clos]]

      def nd(closures: List[Clos]): Set[VS] = {
        val Clos(Lam(v, body), c_env)::rest = closures
        val baddr = allocBind(v, new_time)
        val new_env = c_env + (v -> baddr)
        val new_store = store.update(baddr, aeval(ae, env, store))
        val new_cont = (r: MCont) => (vs: VS) => {
          val baddr = allocBind(x, vs.time)
          val new_env = env + (x -> baddr)
          val new_store = vs.store.update(baddr, vs.vals)
          aval(e, new_env, new_store, vs.time, cont, r)
        }
        val new_rcont = if (rest.isEmpty) {
          rcont 
        } else (vs: VS) => {
          nd(rest) + vs
        }
        aval(body, new_env, new_store, new_time, new_cont, new_rcont)
      }

@@

    case ae if isAtomic(ae) =>
      val ds = aeval(ae, env, store)
      cont(rcont)(VS(ds, new_time, store))
  }
}

@@

\subsubsection{Caching Fixed-Points}

@@

\subsection{Pushdown Control-Flow Analysis}

@@

In this section, we firstly show a simple but uncomputable solution to the
problem; then briefly review several recent research works that enable
computable and precise call/return match. Even though the first solution is
uncomputable, but as we will see in section 3, it is the starting point of our
transformation.

\subsubsection{AAM with Unbound Stack}

A naive solution to is to use precise call stack as we did in concrete
CESK machine. Instead of allocating continuations on store and recording an 
address of continuations in state, we leave the call stack unabstracted.

The state transition function @step@ is still non-deterministic, but
the only non-determinism happens when dereferencing the callee @f@ 
from the store at call-site @App(f, ae)@. 
For each closure of @f@ in the set, we do exactly
the same as what we did in concrete CESK machines: a new frame is constructed
and pushed onto the stack.
In the second case, since the continuation @konts@ is a list of 
frames, we can simply decompose it to differentiate an empty stack case
or a frame exists on the top of stack. 
If the stack is empty, then empty list of successors is returned, 
because the computation is finished.
Otherwise, we matched a frame @Frame(x, e, env_f)@ on the top of stack.
The frame is then poped, and the new state uses the rest of stack.
Since we can only access one frame at a time, this transition is deterministic.

\begin{lstlisting}
def step(s: State): List[State] = {
  val new_time = tick(s)
  s match {
    case State(Let(x, App(f, ae), e), env, bstore, konts, time) if isAtomic(ae) =>
      for (Clos(Lam(v, body), env_c) <- atomicEval(f, env, bstore).toList) yield {
        val frame = Frame(x, e, env)
        val baddr = allocBind(v, new_time)
        val new_env = env_c + (v -> baddr)
        val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
        State(body, new_env, new_store, frame::konts, new_time)
      }
    case State(ae, env, bstore, konts, time) if isAtomic(ae) =>
      konts match {
        case Nil => List()
        case Frame(x, e, env_f)::konts =>
          val baddr = allocBind(x, new_time)
          val new_env = env_f + (x -> baddr)
          val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
          List(State(e, new_env, new_store, konts, new_time))
      }
  }
}
\end{lstlisting}

Unfortunately, even though other components is finite, this AAM with unbounded stack is uncomputable. 
Because the unbounded stack can grow to arbitrary deep and this implies that the state space is possibly 
infinite, therefore the analysis may not terminate for all programs if we simply enumerating 
reachable states.
To see this, consider a program that has two mutually recursived functions:

>>>>

  Following the idea that evaluation contexts are defunctionalized 
  continuations, we also found refunctionalization sequentializes the 
  non-deterministic abstract evaluation.
  Remarkably, we reveal how precise call/return match in control-flow 
  analysis can be obtained by properly refunctionalizing a small-step abstract interpreter.
  As well as we explain how it been lost by defunctionalizing the higher-order
  continuations into an first-order data type.

@@

\subsection{One Step Back: Unabstracted Stack}

In this section, we describe a variant of AAM that leaves stack to be 
unabstracted. That is to to use a precise call stack as we did in concrete CESK machine. 
Instead of allocating continuations on store and embedding
addresses of continuations in states, we intended to use a list of frames
to explicitly model the stack.
By doing so, we recover the call stack as precise as runtime 
(so called pushdown control-flow analysis), but since 
the stack is unbounded thus the analysis is possibly not computable.
For readers who do not familiarized pushdown analysis, we have a detailed 
discussion in Section \ref{pdcfarevisit}.
The reason we do so here is to establish an artifact with precise call/return match
used for the next step of transformation.

The state transition function @step@ is still non-deterministic, but
the only non-determinism happens when dereferencing the callee @f@ 
from function applications @App(f, ae)@.

\begin{lstlisting}
def step(s: State): List[State] = {
  val new_time = tick(s)
  s match {
    case State(Let(x, App(f, ae), e), env, bstore, konts, time) if isAtomic(ae) =>
      for (Clos(Lam(v, body), env_c) <- atomicEval(f, env, bstore).toList) yield {
        val frame = Frame(x, e, env)
        val baddr = allocBind(v, new_time)
        val new_env = env_c + (v -> baddr)
        val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
        State(body, new_env, new_store, frame::konts, new_time)
      }
    case State(ae, env, bstore, konts, time) if isAtomic(ae) =>
      val Frame(x, e, env_f)::konts = konts
      val baddr = allocBind(x, new_time)
      val new_env = env_f + (x -> baddr)
      val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
      List(State(e, new_env, new_store, konts, new_time))
  }
}
\end{lstlisting}

In the first case of state matching, we may have multiple choices of closure for @f@.
For each closure in the set, a new frame is constructed and pushed onto the stack.
The code for handling the second case is the same as concrete CESK machines.
The function @drive@ should be also updated to prevent we pass a state with
empty frame list into function @step@.

\textbf{Note on Computability.}
Unfortunately, even though other components in states are finite, 
this AAM with unbounded stack is still not computable. 
Because the unbounded stack can grow to arbitrary deep which implies that the state space is possibly 
infinite, therefore the analysis may not terminate for all programs if we simply enumerating 
reachable states.
To see this, consider a program that has two mutually recursived functions:

@@

In last section, we showed that by keeping an unabstracted stack in the state 
space we can recover the precise call/return match. In this section, we start
our first transformation: \textit{fusing} an AAM with unbounded-stack.

In fact, as later we will eliminate the stack and transform it to continuations
that are represented by higher-order functions, it does not matter what kind of 
AAM we start from. 
Because of transforming to higher-order functions forces us to sequentialize 
the order of abstract evaluation.
We choose to start from AAM with unbounded-stack just because it has a 
straightforward correspondence of pushdown semantics to refunctionalized AAM.

\begin{lstlisting}
def drive_step(todo: List[State], seen: Set[State]): Set[State] = todo match {
  case Nil => seen
  case hd::tl =>
    if (seen.contains(hd)) return drive_step(tl, seen)

    val new_time = tick(hd)
    val new_states = hd match {
      case State(Let(x, App(f, ae), e), env, bstore, konts, time) 
        if isAtomic(ae) =>
        for (Clos(Lam(v, body), env_c) <- atomicEval(f, env, bstore).toList) yield {
          val frame = Frame(x, e, env)
          val baddr = allocBind(v, new_time)
          val new_env = env_c + (v -> baddr)
          val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
          State(body, new_env, new_store, frame::konts, new_time)
        }
      case State(ae, env, bstore, konts, time) if isAtomic(ae) =>
        konts match {
          case Nil => List()
          case Frame(x, e, f_env)::konts =>
            val baddr = allocBind(x, new_time)
            val new_env = f_env + (x -> baddr)
            val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
            List(State(e, new_env, new_store, konts, new_time))
        }
    }
    drive_step(new_states ++ tl, seen + hd)
}

@@

\begin{lstlisting}
def aval(state: State, todo: List[State], seen: Set[State]): Set[State] = { 
  if (seen.contains(state)) return collect(todo, seen)

  val newTime = tick(state)
  state match {
    case State(Let(x, App(f, ae), e), env, bstore, konts, time) if isAtomic(ae) =>
      val newStates = for (Clos(Lam(v, body), c_env) <- atomicEval(f, env, bstore).toList) yield {
        val frame = Frame(x, e, env)
        val baddr = allocBind(v, newTime)
        val newEnv = c_env + (v -> baddr)
        val newStore = bstore.update(baddr, atomicEval(ae, env, bstore))
        State(body, newEnv, newStore, frame::konts, newTime)
      }   
      collect(newStates ++ todo, seen + state)
    case State(ae, env, bstore, konts, time) if isAtomic(ae) =>
      continue(konts, atomicEval(ae, env, bstore), bstore, newTime, todo, seen + state)
  }   
}

def collect(todo: List[State], seen: Set[State]): Set[State] = { 
  todo match {
    case Nil => seen
    case hd::tl => aval(hd, tl, seen)
  } 
}

@@

As we pointed earlier, refunctionalizing evaluation contexts to higher-order
continuation functions forces us to somehow sequentialize the abstract evaluation.
We identify that in order to sequentialize abstract evaluation, there are three
fundamental differences with concrete evaluation: 
1) The dereferencing store addresses is non-deterministic, therefore the returning
may also be non-deterministic.
2) We need sound collecting semantics rather than valuation semantics.
3) The abstract evaluation has to be terminating.

@@

\subsection{First Try}

@@

def aval(e: Expr, env: Env, store: BStore, time: Time, cont: Cont, rcont: MCont): Set[VS] = {
  val new_time = (e::time).take(k)
  e match {
    case Let(x, App(f, ae), e) =>
      val closures = atomicEval(f, env, store).toList.asInstanceOf[List[Clos]]

      def nd(closures: List[Clos]): Set[VS] = {
        val Clos(Lam(v, body), c_env)::rest = closures
        val baddr = allocBind(v, new_time)
        val new_env = c_env + (v -> baddr)
        val new_store = store.update(baddr, atomicEval(ae, env, store))
        val new_cont = (r: MCont) => (vs: VS) => {
          val baddr = allocBind(x, vs.time)
          val new_env = env + (x -> baddr)
          val new_store = vs.store.update(baddr, vs.vals)
          aval(e, new_env, new_store, vs.time, cont, r)
        }
        val new_rcont = if (rest.isEmpty) {
          rcont 
        } else (vs: VS) => {
          nd(rest) + vs
        }
        aval(body, new_env, new_store, new_time, new_cont, new_rcont)
      }

@@

    case ae if isAtomic(ae) =>
      val ds = atomicEval(ae, env, store)
      cont(rcont)(VS(ds, new_time, store))
  }
}

@@

\subsection{Caching Fixed-Points}

@@

\subsection{Pushdown Control-Flow Analysis, Revisited} \label{pdcfarevisit}

@@
